\documentclass[11pt]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{parskip}
\usepackage{xcolor}

\title{Why We Threw Away the Cascade Controller\\
       \large A Chapter 4 Architecture Decision Note}
\author{Hari Vasudevan}
\date{\today}

\begin{document}
\maketitle

\textit{Chapters 1 through 3 of this series applied classical cascade
control to LLM inference serving.  It was elegant on paper and worked
reasonably well in simulation.  Then we connected it to a real GPU and it
told us something we did not expect: the plant we had been designing for
does not exist.  This note documents what we found, why it happened, and
what we built instead.}

% ─────────────────────────────────────────────────────────────────────────────
\section*{Where We Started}
% ─────────────────────────────────────────────────────────────────────────────

The core assumption running through Chapters 1--3 was that an LLM inference
server behaves like a service queue.  Requests arrive at rate $\lambda$,
the server processes them in batches of size $B$, and any excess builds up
as a pending queue of depth $q$.  Under that assumption, the p95 latency is
well-described by:

\begin{equation}
    \ell[k] = \alpha B[k] + \gamma B[k]^2 + \beta\, q[k]
    \label{eq:model}
\end{equation}

The three terms capture service time (linear and quadratic in batch size)
and queueing delay (linear in queue depth).  We set $\alpha = 0.1$,
$\gamma = 0.8$, and $\beta = 2.0\;\text{ms/req}$ from a theoretical
stochastic simulation, and designed a two-loop cascade controller around
this model.  The inner loop regulated $q$ by adjusting $B$, and the outer
loop regulated $\ell$ by setting the queue setpoint $q_{\mathrm{ref}}$.

The model was reasonable.  The problem was that it described a different
system than the one we were actually running against.

% ─────────────────────────────────────────────────────────────────────────────
\section*{What the Real System Told Us}
% ─────────────────────────────────────────────────────────────────────────────

When we connected the Chapter~3 cascade controller to a live
\texttt{qwen2.5:3b} model served by Ollama on an M2 MacBook Air, two things
became apparent almost immediately.

\subsection*{The queue never builds}

We ran a plant identification experiment: sustained Poisson arrivals at
five different rates $\lambda \in \{2, 3, 5, 7, 9\}$ requests per tick,
each run for 35 ticks.  The drain rule was straightforward --- set
$B[k] = \min(B_{\max},\, q[k] + a[k])$ so the server tries to absorb
everything it receives.  We expected to see different steady-state queue
depths at different lambda values.  Instead, we saw $q \approx 0$ in every
single run, regardless of load.

The reason is structural.  Unlike a production serving engine such as vLLM,
which maintains an explicit scheduler queue and can hold requests pending
while others are processed, Ollama dispatches each incoming HTTP request
immediately to an available inference slot.  When all slots are occupied,
requests wait inside Ollama's internal HTTP stack --- not in a counter that
MATLAB can observe.  From our software queue's perspective, arrivals were
always matched immediately.  The $\beta q$ term in
equation~\eqref{eq:model} has no observable counterpart in this system.

This is not a bug in our experiment.  It is a fundamental property of how
Ollama is architected.  Chapter~4 will be honest about this distinction:
Ollama is a development tool, not a production serving engine.  The right
conclusions here carry over to vLLM, which does expose true queue depth ---
but that is Chapter~5's problem.

\subsection*{Latency is a function of concurrency, not queue depth}

Having established that $q \approx 0$ always, we re-ran the identification
with a different question: how does TTFT (Time to First Token) change as we
fire more requests \emph{simultaneously}?  We swept $B \in \{1, \ldots, 12\}$
by submitting $B$ requests concurrently via \texttt{parfeval}, measured the
mean TTFT across five repetitions at each point, and fitted a quadratic:

\begin{equation}
    \ell(B) = c_0 + c_1 B + c_2 B^2
    \label{eq:fit}
\end{equation}

The fit came back:
\begin{align*}
    c_0 &= 97.16 \;\text{ms} \qquad \text{(prefill cost with no contention)} \\
    c_1 &= 30.35 \;\text{ms/req} \qquad \text{(linear contention growth)} \\
    c_2 &= -1.44 \;\text{ms/req}^2 \qquad \text{(concavity at high $B$)}
\end{align*}

with $R^2 = 0.815$.  The physical story here is GPU KV-cache memory
bandwidth competition.  When $B$ sequences are processed in parallel, they
compete for the same memory bus, and prefill time rises.  The negative $c_2$
reflects the fact that at high concurrency Ollama spreads load across
multiple inference slots (we had configured
\texttt{OLLAMA\_NUM\_PARALLEL=16}), partially offsetting the linear term.
This is a concurrency effect, not a queueing effect.

\subsection*{Our assumed $\beta$ was off by $8\times$}

This is where the cascade controller design unravelled.  Linearising
equation~\eqref{eq:fit} at our chosen operating point $B_0 = 5$:

\begin{equation}
    \beta_{\mathrm{eff}}
      = \left.\frac{d\ell}{dB}\right|_{B_0}
      = c_1 + 2c_2 B_0
      = 30.35 - 2(1.44)(5)
      = 15.97 \;\text{ms/req}
\end{equation}

We had been designing the outer loop with $\beta = 2.0\;\text{ms/req}$.
The real sensitivity was $15.97\;\text{ms/req}$ --- an eightfold difference.
Since the integral gain is $K_{il} = (e^{-\Delta t/\tau} - 1)/\beta$, we
had been running a controller that was $8\times$ more aggressive than the
plant required.  The integrator was over-correcting on every tick, which
explains the oscillatory behaviour we observed in the Chapter~3 runs.

% ─────────────────────────────────────────────────────────────────────────────
\section*{The Revised Architecture}
% ─────────────────────────────────────────────────────────────────────────────

Once we accepted that $q \approx 0$ always and the plant reduces to the
static map $\ell \approx f(B)$, the case for the cascade evaporated.  The
inner loop had been regulating a state that does not exist in the real
system.  The outer loop was commanding a queue setpoint that always resolved
to zero.  The complexity was not earning us anything.

The Chapter~4 controller is a single-state discrete-time integrator that
observes TTFT and adjusts $B$ directly:

\begin{align}
    e[k]      &= \ell_{\mathrm{target}} - \ell_{\mathrm{meas}}[k] \\
    \xi[k+1]  &= \xi[k] + e[k] \label{eq:int} \\
    B[k]      &= \mathrm{clamp}\!\left(B_0 - K_{il}\,\xi[k],\;
                                       B_{\min},\; B_{\max}\right)
\end{align}

The closed-loop characteristic equation follows from substituting the
control law back into the integrator update and using $\ell = \beta_{\mathrm{eff}} \Delta B$:

\begin{equation}
    z_{\mathrm{cl}} = 1 + \beta_{\mathrm{eff}}\, K_{il}
\end{equation}

We want this pole at $z_{\mathrm{cl}} = e^{-\Delta t / \tau_{\mathrm{cl}}}$,
which gives the gain analytically:

\begin{equation}
    K_{il} = \frac{e^{-\Delta t/\tau_{\mathrm{cl}}} - 1}{\beta_{\mathrm{eff}}}
           = \frac{e^{-1/20} - 1}{15.97}
           = -0.003054 \;\text{req/ms}
\end{equation}

We chose $\tau_{\mathrm{cl}} = 20\;\text{s}$ to ensure the closed-loop
bandwidth sits comfortably below the measurement lag introduced by the
20-sample rolling window used to compute $\ell_{\mathrm{meas}}$.  Pushing
the controller faster than the measurement filter is a reliable recipe for
integrator windup, and we had already learned that lesson in Chapter~3.

The anti-windup bounds on the integrator state $\xi$ are derived directly
from the actuator limits.  Since $B = B_0 - K_{il}\xi$ and
$B \in [B_{\min}, B_{\max}]$, the admissible range for $\xi$ is:

\begin{equation}
    \xi \in \left[\frac{B_0 - B_{\max}}{K_{il}},\;
                   \frac{B_0 - B_{\min}}{K_{il}}\right]
         = [-1310,\; 2292]
\end{equation}

These are wide bounds by design.  We have only one integrator now, and
we want it to have full authority over the $B$ range $[1, 12]$ without
unnecessary clamping during normal operation.

% ─────────────────────────────────────────────────────────────────────────────
\section*{What Changed and Why It Matters}
% ─────────────────────────────────────────────────────────────────────────────

The table below summarises the difference between the Chapter~3 and
Chapter~4 architectures.  The number of controller states halved.  The
assumed $\beta$ was replaced by a measured one.  The resulting gain is
$5.4\times$ smaller in magnitude, which matches what the real GPU-memory
dynamics actually require.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{p{3.5cm} p{5.5cm} p{5.5cm}}
\toprule
& \textbf{Chapter 3 (cascade)} & \textbf{Chapter 4 (single-loop)} \\
\midrule
Plant model
  & $\ell = \alpha B + \gamma B^2 + \beta q$ (assumed)
  & $\ell = c_0 + c_1 B + c_2 B^2$ (measured) \\
Controller states
  & $\xi_q$ (inner) $+$ $\xi_l$ (outer)
  & $\xi$ (one integrator) \\
Actuator path
  & $\xi_l \to q_{\mathrm{ref}} \to \xi_q \to B$
  & $\xi \to B$ directly \\
$\beta$ used
  & 2.0 ms/req (from simulation)
  & 15.97 ms/req (from identification) \\
$K_{il}$
  & $-0.0164$ ($8\times$ too aggressive)
  & $-0.003054$ (matched to plant) \\
Queue state
  & Tracked in software, used as control state
  & Not needed ($q \approx 0$ always in Ollama) \\
\bottomrule
\end{tabular}
\caption{Architecture comparison between Chapter~3 and Chapter~4.}
\end{table}

% ─────────────────────────────────────────────────────────────────────────────
\section*{What This Means for Chapter 5 (vLLM)}
% ─────────────────────────────────────────────────────────────────────────────

The most important takeaway from this exercise is not the gain value or the
controller simplification --- it is the distinction between Ollama and a
production serving engine.  In vLLM, the queue depth \texttt{num\_requests\_waiting}
is a real, observable, controllable state exposed via the Prometheus metrics
endpoint.  The cascade architecture from Chapters 2b and 3 is not wrong; it
is just solving the right problem for the wrong plant.

Chapter~5 will reconnect the inner queue-regulation loop.  The outer loop
integral gain will need to be re-identified against vLLM's actual response
surface, which includes true scheduler-level queueing, preemption, and
continuous batching dynamics that have no analogue in Ollama.  We expect
$\beta$ to be much closer to our original assumed value of 2~ms/req in that
setting, because the $\beta q$ term in equation~\eqref{eq:model} will
finally have a real plant to act on.

The architectural simplification in Chapter~4 is therefore a deliberate
choice, not a permanent one.  We are building the simplest correct
controller for the plant we actually have, rather than forcing a more
complex architecture onto a system that cannot support it.

\end{document}
